---
output: html_document
---
Factors Affecting White Wine Quality by Caleb Sangwin
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
library(gridExtra)
library(corrplot)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
# Load the Data
wwq <- read.csv('wineQualityWhites.csv')
wwq <- wwq[, c(2:13)] # Removing 'X' the column for observation ID.

# Changing sulfur dioxide into g / dm^3 like the rest of the variables
wwq$free.sulfur.dioxide <- wwq$free.sulfur.dioxide / 1000
wwq$total.sulfur.dioxide <- wwq$total.sulfur.dioxide / 1000
wwq$used.sulfur.dioxide <- wwq$total.sulfur.dioxide - wwq$free.sulfur.dioxide

# Creating main code up here to keep next section tidy
# Creating high and low quality wine datasets
highquality <- subset(wwq, quality > 6)
lowquality <- subset(wwq, quality < 6)

# Storing summaries as variables
all.data.summary <- summary(wwq) # dataset with all wines
high.quality.summary <- summary(highquality) # dataset with wines rated 7+
low.quality.summary <- summary(lowquality) # dataset with wines rated 5-
# Removing {Min, 1st Q, Median, Mean, 3rd Q, Max} labels from the summary data
# Unsure how to do this programmatically in R
all.data.summary <- substr(all.data.summary, 9, 14) 
high.quality.summary <- substr(high.quality.summary, 9, 14) 
low.quality.summary <- substr(low.quality.summary, 9, 14) 
# Names for rows and columns
summaryrows <- c('Min', '1st Q', 'Median', 'Mean', '3rd Q', 'Max')
summarycolumns <- c('Low Quality', 'All Wine', 'High Quality')
# A function creating a table to show summary values across three datasets
summarytable <- function(col_id) {
  new.matrix <- matrix(c(low.quality.summary[,col_id],
                         all.data.summary[,col_id],
                         high.quality.summary[,col_id]), 
                      ncol = 3)
  colnames(new.matrix) <- summarycolumns
  rownames(new.matrix) <- summaryrows
  new.matrix
}

# Quick view of the Data Structure
str(wwq)
```

> This data set on white wine quality has 4898 observations of 13 variables.
Quality is the output observation and is the only integer in this dataset. The 
rest of the variables are decimal point numbers. I want to see if quality can 
be estimated based on any of these variables.

# Univariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_Quality}
# Summary of quality
print('Quality')
summary(wwq$quality)

# Number of high quality wines
print('High Quality Wines')
nrow(highquality) 

# Number of low quality wines
print('Low Quality Wines')
nrow(lowquality) 

# Histogram of quality of the whole dataset
ggplot(aes(x = quality), data = wwq) + 
  geom_bar()
```

> I grouped the wines rated 7 or greater and grouped the wines rated 5 or 
lower. Close to 21% of the wines are high quality and 33% are low quality. I'm
hoping to find similarities in these groups and compare them to each other
and the dataset as a whole. The histogram showing the quality of the white 
wines has a fairly normal distribution. 

### Acidity

```{r echo=FALSE, message=FALSE, warning=FALSE, Acidity_1}
# Fixed Acidity Histogram and Box Plot
fa.h <- ggplot(aes(x = fixed.acidity), data = wwq) + 
  geom_histogram(binwidth = 0.1)
fa.b <- ggplot(aes(x = "", y = fixed.acidity), data = wwq) +
  geom_boxplot(color = 'darkred') +
  geom_jitter(alpha = 0.1)
# Volatile Acidity Histogram and Box Plot
va.h <- ggplot(aes(x = volatile.acidity), data = wwq) + 
  geom_histogram(binwidth = 0.01)
va.b <- ggplot(aes(x = "", y = volatile.acidity), data = wwq) +
  geom_boxplot(color = 'darkred') +
  geom_jitter(alpha = 0.1)

# Fixed Acidity Summary and Plots
print('Fixed Acidity')
summary(wwq$fixed.acidity)
grid.arrange(fa.h, fa.b, ncol = 2)
# Volatile Acidity Summary and Plots
print('Volatile Acidity')
summary(wwq$volatile.acidity)
grid.arrange(va.h, va.b, ncol = 2)
```

> Both Fixed and Volatile Acidity have fairly normal distributions. Volatile
Acidity has many more outliers, but both of these variales have small
interquartile ranges. It's interesting that the IQR makes up about 10% of the
range for both of these variables. This seems very low and speaks to the 
variability in this dataset.

```{r echo=FALSE, message=FALSE, warning=FALSE, Acidity_2}
# Citric Acid Histogram and Box Plot
ca.h <- ggplot(aes(x = citric.acid), data = wwq) + 
  geom_histogram(binwidth = 0.01) +
  scale_x_continuous(limits = c(0, 1))
ca.b <- ggplot(aes(x = "", y = citric.acid), data = wwq) +
  geom_boxplot(color = 'darkred') +
  geom_jitter(alpha = 0.05)
# pH Histogram and Box Plot
ph.h <- ggplot(aes(x = pH), data = wwq) + 
  geom_histogram(binwidth = 0.01)
ph.b <- ggplot(aes(x = "", y = pH), data = wwq) +
  geom_boxplot(color = 'darkred') +
  geom_jitter(alpha = 0.1)

# Citric Acid Summary and Plots
print('Citric Acid')
summary(wwq$citric.acid)
grid.arrange(ca.h, ca.b, ncol = 2)
# pH Summary and Plots
print('pH')
summary(wwq$pH)
grid.arrange(ph.h, ph.b, ncol = 2)
```

> Citric Acid and pH again have fairly normal distributions with quite a few
outliers. This time though Citric Acid has an IQR that makes up only 7% of
the range while pH is closer to 17%. I wonder if high quality wines have
values that fall within these small IQR's. 

### Sulphates and Sulfur Dioxide

```{r echo=FALSE, message=FALSE, warning=FALSE, Sulphates_and_Sulfur_Dioxide}
# Sulphates Histogram and Box Plot
su.h <- ggplot(aes(x = sulphates), data = wwq) + 
  geom_histogram(binwidth = 0.01)
su.b <- ggplot(aes(x = "", y = sulphates), data = wwq) +
  geom_boxplot(color = 'darkred') +
  geom_jitter(alpha = 0.1)
# Free Sulfur Dioxide Histogram and Box Plot
# Limits and binwidth set to improve visual aid across three graphs
fsd.h <- ggplot(aes(x = free.sulfur.dioxide), data = wwq) + 
  geom_histogram(binwidth = 0.001) +
  scale_x_continuous(limits = c(0, 0.3)) +
  scale_y_continuous(limits = c(0, 170))
fsd.b <- ggplot(aes(x = "", y = free.sulfur.dioxide), data = wwq) +
  geom_boxplot(color = 'darkred') +
  geom_jitter(alpha = 0.05) +
  coord_cartesian(ylim = c(0, 0.3))
# Used Sulfur Dioxide Histogram and Box Plot
usd.h <- ggplot(aes(x = used.sulfur.dioxide), data = wwq) + 
  geom_histogram(binwidth = 0.001) +
  scale_x_continuous(limits = c(0, 0.3)) +
  scale_y_continuous(limits = c(0, 170))
usd.b <- ggplot(aes(x = "", y = used.sulfur.dioxide), data = wwq) +
  geom_boxplot(color = 'darkred') +
  geom_jitter(alpha = 0.05) +
  coord_cartesian(ylim = c(0, 0.3))
# Total Sulfur Dioxide Histogram and Box Plot
tsd.h <- ggplot(aes(x = total.sulfur.dioxide), data = wwq) + 
  geom_histogram(binwidth = 0.001) +
  scale_x_continuous(limits = c(0, 0.3)) +
  scale_y_continuous(limits = c(0, 170))
tsd.b <- ggplot(aes(x = "", y = total.sulfur.dioxide), data = wwq) +
  geom_boxplot(color = 'darkred') +
  geom_jitter(alpha = 0.05) +
  coord_cartesian(ylim = c(0, 0.3))

# Sulphates Summary and Plots
print('Sulphates')
summary(wwq$sulphates)
grid.arrange(su.h, su.b, ncol = 2)
# Sulfur Dioxide Plots
print('Sulfur Dioxide')
grid.arrange(fsd.h, usd.h, tsd.h, ncol = 3)
grid.arrange(fsd.b, usd.b, tsd.b, ncol = 3)
# Sulfur Dioxide Summaries
print('Free Sulfur Dioxide')
summary(wwq$free.sulfur.dioxide)
print('Used Sulfur Dioxide')
summary(wwq$used.sulfur.dioxide)
print('Total Sulfur Dioxide')
summary(wwq$total.sulfur.dioxide)
```

> So far very little has stood out as abnormal in these variables. Sulphates
and Sulfur Dioxide levels are no different, with normal distributions and a
similar number of outliers compared to the other variables. Free Sulfur Dioxide
is the only variable with a small IQR making up about 8% of the full range. The
others have much larger IQR's. 

### Chlorides, Sugar, Density, and Alcohol

```{r echo=FALSE, message=FALSE, warning=FALSE, Chlorides_and_Sugar}
# Chlorides Histogram and Box Plot
ch.h <- ggplot(aes(x = chlorides), data = wwq) + 
  geom_histogram(binwidth = 0.001) +
  scale_x_continuous(limits = c(0, 0.1))
ch.b <- ggplot(aes(x = "", y = chlorides), data = wwq) +
  geom_boxplot(color = 'darkred') +
  geom_jitter(alpha = 0.05)
# Residual Sugar Histogram and Box Plot
rs.h <- ggplot(aes(x = residual.sugar), data = wwq) + 
  geom_histogram(binwidth = 0.2) +
  scale_x_continuous(limits = c(0, 20))
rs.b <- ggplot(aes(x = "", y = residual.sugar), data = wwq) +
  geom_boxplot(color = 'darkred') +
  geom_jitter(alpha = 0.1) +
  coord_cartesian(ylim = c(0, 20))

# Chloride Summary and Plots
print('Chlorides')
summary(wwq$chlorides)
grid.arrange(ch.h, ch.b, ncol = 2)
# Residual Sugar Summary and Plots
print('Residual Sugar')
summary(wwq$residual.sugar)
grid.arrange(rs.h, rs.b, ncol = 2)
```

> These two variables are much different. Chlorides has a fairly normal
distribution with a bit of a positive skew. It has the smallest IQR when 
compared to range at about 4%. Residual Sugar on the other hand is very
positively skewed, but has a large amount of data at low levels. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Density_and_Alcohol}
# Density Histogram and Box Plot
de.h <- ggplot(aes(x = density), data = wwq) + 
  geom_histogram(binwidth = 0.0002) +
  scale_x_continuous(limits = c(0.987, 1.005))
de.b <- ggplot(aes(x = "", y = density), data = wwq) +
  geom_boxplot(color = 'darkred') +
  geom_jitter(alpha = 0.1) +
  coord_cartesian(ylim = c(0.987, 1.005))
# Alcohol Histogram and Box Plot
al.h <- ggplot(aes(x = alcohol), data = wwq) + 
  geom_histogram(binwidth = 0.1)
al.b <- ggplot(aes(x = "", y = alcohol), data = wwq) +
  geom_boxplot(color = 'darkred') +
  geom_jitter(alpha = 0.1)

# Density Summary and Plots
print('Density')
summary(wwq$density)
grid.arrange(de.h, de.b, ncol = 2)
# Alcohol Summary and Plots
print('Alcohol')
summary(wwq$alcohol)
grid.arrange(al.h, al.b, ncol = 2)
```

> For these last two variables they are again distributed quite normally, with
the exception that alcohol has a bit of a positive skew. Nothing else really 
stands out.

# Univariate Analysis

### What is the structure of your dataset?

> This dataset contains 4989 observations of 13 variables. All of the 12 
input variables are decimal point numbers and the one output variable is
an integer. With the exception of sugar, all of the variables are found in 
very small quantities - less than 1 g / dm^3. 

> Quality ranges from 3 to 9 with a median of 6 and a mean of 5.878. 

### What is/are the main feature(s) of interest in your dataset?

> The main feature of interest is our output variable - quality. The goal
is to determine which is the measured variables affect the overall
quality. The attribute details state that the output is based on sensory
data which I assume to mean quality is based on taste and smell of the
wine (more so than hangover level or overall price value). 

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

> After doing a little research to determine what affects the taste of wine,
it appears that acidity and sulfur dioxide levels may be a good indicator.
Acidity causes food and drinks to taste more zesty. Citric acid may add
to the sweetness but fixed and volatile acidity will add more zest. During
the fermentation process, maintaining appropriate oxygen levels is important
as oxygen causes reactions in the other chemicals in the wine. This reaction
causes the fruit in the wine to lose it's aroma. Sulfur Dioxide is added as an 
antioxidant to help reduce reactions between molecules.

> From what I saw above, Chlorides, Alcohol, and Sugar levels were the most
different of all the variables. I think they would be good places to start. 

### Did you create any new variables from existing variables in the dataset?

> I created a variable for the used sulfur dioxide by taking the total
sulfur dioxide and subtracting the free sulfur dioxide in each wine. I did 
this because sulfur dioxide may be useful in determining quality, however
I'm unsure whether available or used sulfur dioxide has more affect. 

> I also changed the values for the sulfur dioxide variables so that they
were measured in the same units as the rest of the variables. And I removed
the variable 'X' which was just an observation ID. 

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

> None of the distributions jumped out as unusual. There are however a lot of
outliers in nearly each variable. The only change I made was to create datasets 
containing the best and worst quality wines. I think finding relationships 
between wines of similar quality will be more appropriate than trying to 
determine the same thing over an entire group full of outliers. 

# Bivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Correlation}
# Storing correlation dataframe in variables to use in corrplot
wwq.cor <- cor(wwq) 
corrplot(wwq.cor, method = 'circle')
```

> Here is the correlation plot for White Wine. The most correlated
variable with quality appears to be alcohol with a value of 0.43. This will be 
a good start for our analysis. Alcohol also appears to be at least partially 
correlated with sugar, chlorides, total sulfur dioxide, and density.

> Density has two of the highest correlation values when compared to sugar and
alcohol with values of 0.84 and -0.78 respectively. I think these two directly 
affect the density of wine and won't be used towards the quality. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Quality_vs_Alcohol}
# Histogram showing quality vs alcohol and Correlation
ggplot(aes(x = quality, y = alcohol), data = wwq) +
  geom_jitter(alpha = 0.25) +
  geom_smooth(method = 'lm', se = FALSE)
print('Correlation between Quality and Alcohol')
cor(wwq$quality, wwq$alcohol)
# Alcohol vs Quality Box Plot and Jitter
ggplot() +
  geom_boxplot(aes(x = "3 - High", y = alcohol), # High Quality Box Plot
               data = highquality, color = 'darkred') + 
  geom_jitter(aes(x = '3 - High', y = alcohol), # High Quality Jitter
              data = highquality, alpha = 0.1) + 
  geom_boxplot(aes(x = "2 - All", y = alcohol), # All Data Box Plot
               data = wwq, color = 'darkred') +
  geom_jitter(aes(x = '2 - All', y = alcohol), # All Data Jitter
              data = wwq, alpha = 0.1) + 
  geom_boxplot(aes(x = "1 - Low", y = alcohol), # Low Quality Box Plot
               data = lowquality, color = 'darkred') +
  geom_jitter(aes(x = '1 - Low', y = alcohol), # Low Quality Jitter
              data = lowquality, alpha = 0.1) + 
  xlab('Quality')

# Alcohol vs Quality Summary
alcohol.summary <- summarytable(11) # Won't work when called with name?
print('alcohol')
alcohol.summary
```

> The most correlated variable with quality is alcohol. I like using 
geom_jitter with an alpha value because it's a little easier to understand 
visually. There is definitely an upward trend seen between quality and
alcohol. High quality wines generally have higher alcohol contents when
compared to the low quality wines.

```{r echo=FALSE, message=FALSE, warning=FALSE, Density_vs_Alcohol/Sugar}
# Residual Sugar vs Quality Box Plot and Jitter
rs.b <- ggplot() +
  geom_boxplot(aes(x = "3 - High", y = residual.sugar), # High Quality Box Plot
               data = highquality, color = 'darkred') + 
  geom_jitter(aes(x = '3 - High', y = residual.sugar), # High Quality Jitter
              data = highquality, alpha = 0.1) + 
  geom_boxplot(aes(x = "2 - All", y = residual.sugar), # All Data Box Plot
               data = wwq, color = 'darkred') + 
  geom_jitter(aes(x = '2 - All', y = residual.sugar), # All Data Jitter
              data = wwq, alpha = 0.1) + 
  geom_boxplot(aes(x = "1 - Low", y = residual.sugar), # Low Quality Box Plot
               data = lowquality, color = 'darkred') + 
  geom_jitter(aes(x = '1 - Low', y = residual.sugar), # Low Quality Jitter
              data = lowquality, alpha = 0.1) + 
  coord_cartesian(ylim = c(0, 20)) + 
  xlab('Quality')
# Density vs Quality Box Plot and Jitter
de.b <- ggplot() +
  geom_boxplot(aes(x = "3 - High", y = density), # High Quality Box Plot
               data = highquality, color = 'darkred') + 
  geom_jitter(aes(x = '3 - High', y = density), # High Quality Jitter
              data = highquality, alpha = 0.1) + 
  geom_boxplot(aes(x = "2 - All", y = density), # All Data Box Plot
               data = wwq, color = 'darkred') +
  geom_jitter(aes(x = '2 - All', y = density), # All Data Jitter
              data = wwq, alpha = 0.1) + 
  geom_boxplot(aes(x = "1 - Low", y = density), # Low Quality Box Plot
               data = lowquality, color = 'darkred') +
  geom_jitter(aes(x = '1 - Low', y = density), # Low Quality Jitter
              data = lowquality, alpha = 0.1) + 
  coord_cartesian(ylim = c(0.989, 1)) + 
  xlab('Quality')
# Density vs Residual Sugar
d.rs <- ggplot(aes(x = density, y = residual.sugar), data = wwq) +
  geom_point(alpha = 1/10) +
  coord_cartesian(xlim = c(0.985, 1.005), ylim = c(0, 30)) +
  geom_smooth(method = 'lm', se = FALSE)
# Density vs Alcohol
d.al <- ggplot(aes(x = density, y = alcohol), data = wwq) +
  geom_point(alpha = 1/10) +
  coord_cartesian(xlim = c(0.985, 1.005), ylim = c(7.5, 15)) +
  geom_smooth(method = 'lm', se = FALSE)
# Density and Sugar vs Quality Summaries
density.summary <- summarytable(8) # Won't work when called with name?
sugar.summary <- summarytable('residual.sugar')

# Density and Sugar vs Quality Box Plots and Summaries
grid.arrange(rs.b, de.b, ncol = 2)
print('Sugar')
sugar.summary
print('Density')
density.summary
# Sugar and Alcohol vs Density Plots and Correlation
grid.arrange(d.rs, d.al, ncol = 2)
print('Correlation between Density and Sugar')
cor(wwq$density, wwq$residual.sugar)
print('Correlation between Density and Alcohol')
cor(wwq$density, wwq$alcohol)
```

> The box plots clearly show the change of a few key measurements between high
and low quality wine. High quality wines generally have a smaller IQR and range.
This is similar to what was found before when looking at single variables. 

> Since density is so heavily correlated to alcohol and sugar, it won't be used
to determine quality. Sugar and alcohol are qualities that are more likely to 
have a direct affect. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Alcohol_vs_Chlorides/TSD/Sugar}
# Chlorides vs Quality Box Plot and Jitter
ch.b <- ggplot() +
  geom_boxplot(aes(x = "3 - High", y = chlorides), # High Quality Box Plot
               data = highquality, color = 'darkred') + 
  geom_jitter(aes(x = '3 - High', y = chlorides), # High Quality Jitter
              data = highquality, alpha = 0.05) + 
  geom_boxplot(aes(x = "2 - All", y = chlorides), # All Data Box Plot
               data = wwq, color = 'darkred') + 
  geom_jitter(aes(x = '2 - All', y = chlorides), # All Data Jitter
              data = wwq, alpha = 0.05) + 
  geom_boxplot(aes(x = "1 - Low", y = chlorides), # Low Quality Box Plot
               data = lowquality, color = 'darkred') + 
  geom_jitter(aes(x = '1 - Low', y = chlorides), # Low Quality Jitter
              data = lowquality, alpha = 0.05) + 
  coord_cartesian(ylim = c(0, 0.1)) + 
  xlab('Quality')
# Total Sulfur Dioxide vs Quality Box Plot and Jitter
tsd.b <- ggplot() +
  geom_boxplot(aes(x = "3 - High", y = total.sulfur.dioxide), # High Quality Box Plot
               data = highquality, color = 'darkred') + 
  geom_jitter(aes(x = '3 - High', y = total.sulfur.dioxide), # High Quality Jitter
              data = highquality, alpha = 0.05) + 
  geom_boxplot(aes(x = "2 - All", y = total.sulfur.dioxide), # All Data Box Plot
               data = wwq, color = 'darkred') +
  geom_jitter(aes(x = '2 - All', y = total.sulfur.dioxide), # All Data Jitter
              data = wwq, alpha = 0.05) + 
  geom_boxplot(aes(x = "1 - Low", y = total.sulfur.dioxide), # Low Quality Box Plot
               data = lowquality, color = 'darkred') +
  geom_jitter(aes(x = '1 - Low', y = total.sulfur.dioxide), # Low Quality Jitter
              data = lowquality, alpha = 0.05) + 
  coord_cartesian(ylim = c(0, 0.3)) + 
  xlab('Quality')
# Residual Sugar vs Quality Box Plot and Jitter
rs.b <- ggplot() +
  geom_boxplot(aes(x = "3 - High", y = residual.sugar), # High Quality Box Plot
               data = highquality, color = 'darkred') + 
  geom_jitter(aes(x = '3 - High', y = residual.sugar), # High Quality Jitter
              data = highquality, alpha = 0.05) + 
  geom_boxplot(aes(x = "2 - All", y = residual.sugar), # All Data Box Plot
               data = wwq, color = 'darkred') +
  geom_jitter(aes(x = '2 - All', y = residual.sugar), # All Data Jitter
              data = wwq, alpha = 0.05) + 
  geom_boxplot(aes(x = "1 - Low", y = residual.sugar), # Low Quality Box Plot
               data = lowquality, color = 'darkred') +
  geom_jitter(aes(x = '1 - Low', y = residual.sugar), # Low Quality Jitter
              data = lowquality, alpha = 0.05) + 
  coord_cartesian(ylim = c(0, 20)) + 
  xlab('Quality')
# Alcohol vs Chlorides Plot
al.ch <- ggplot(aes(x = alcohol, y = chlorides), data = wwq) +
  geom_point(alpha = 0.1) +
  coord_cartesian(ylim = c(0, 0.1)) +
  geom_smooth(method = 'lm', se = FALSE)
# Alcohol vs Total Sulfur Dioxide Plot
al.tsd <- ggplot(aes(x = alcohol, y = total.sulfur.dioxide), data = wwq) +
  geom_point(alpha = 0.1) +
  coord_cartesian(ylim = c(0, 0.3)) +
  geom_smooth(method = 'lm', se = FALSE)
# Alcohol vs Residual Sugar Plot
al.rs <- ggplot(aes(x = alcohol, y = residual.sugar), data = wwq) +
  geom_point(alpha = 0.1) +
  coord_cartesian(ylim = c(0, 20)) +
  geom_smooth(method = 'lm', se = FALSE)
# Chlorides and Total Sulfur Dioxide Summaries
chlorides.summary <- summarytable(5) # Won't work when called with name?
tsd.summary <- summarytable('total.sulfur.dioxide')

# Chlorides Plots and Summary
grid.arrange(ch.b, al.ch, ncol = 2)
print('Chlorides')
chlorides.summary
print('Correlation')
cor(wwq$alcohol, wwq$chlorides)
# Total Sulfur Dioxide Plots and Summary
grid.arrange(tsd.b, al.tsd, ncol = 2)
print('Total Sulfur Dioxide')
tsd.summary
print('Correlation')
cor(wwq$alcohol, wwq$total.sulfur.dioxide)
# Residual Sugar Plots
grid.arrange(rs.b, al.rs, ncol = 2)
print('Correlation')
cor(wwq$alcohol, wwq$residual.sugar)
```

> Chlorides in high and low quality wine actually have the same sized IQR but
low quality wine has many more outliers. A specific chloride range isn't
enough to determine quality but it definitely plays a role. We already saw
that high quality wines tend to have a high alcohol content, making the
correlation between chlorides and alcohol very helpful. 

> The same can be said for the total sulfur dioxide and sugar plots, but they
have slightly less change between quality levels. Either way it appears that 
these variables are the most likely to affect the quality level.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

> The biggest factor I've seen so far in determining quality level is alcohol.
I didn't expect wine quality to be most correlated with alcohol. Comparing
alcohol against other variables is offering insights into secondary or tertiary
variables that affect quality. Other than alcohol, the three most useful
variables are chlorides, total sulfur dioxide, and sugar. Seeing the range in
the box plots change between quality levels was very useful in selecting
useful variables.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

> I checked to make sure that density levels were caused by other variiables
more than quality. Density is indeed very correlated with both sugar and 
alcohol. This was important to check because density was one of the more
correlated variables with quality when looking at the whole dataset.

### What was the strongest relationship you found?

> The strongest relationship is definitely between alcohol and chlorides,
followed by alcohol and total sulfur dioxide. I think I'll look closer at these
variables to see what else I can determine. 

# Multivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, High_and_Low_Quality_Corr_Plot}
# Storing correlation dataframe in variables to use in corrplot
hq.cor <- cor(highquality) 
lq.cor <- cor(lowquality)
# Easy to compare correlation grids
print('High Quality')
corrplot(hq.cor, method = 'circle')
print('Low Quality')
corrplot(lq.cor, method = 'circle')
```

> Looking at the high quality correlation plot, there are three major 
differences I see when comparing against alcohol. Fixed Acidity, Volatile 
Acidity, and Chlorides all increase greatly. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Alcohol_vs_Chlorides_vs_Quality}
# Alcohol vs Chlorides vs Quality
ggplot() +
  geom_point(aes(x = alcohol, y = chlorides, color = 'High Quality'), 
             data = highquality, alpha = 0.33) +
  geom_point(aes(x = alcohol, y = chlorides, color = 'Low Quality'), 
             data = lowquality, alpha = 0.21) +
  scale_color_manual("",
                     values = c("High Quality"="darkorange4", 
                                "Low Quality"="blue")) +
  geom_smooth(aes(x = alcohol, y = chlorides, color = 'High Quality'),
              data = highquality, se = FALSE) +
  geom_smooth(aes(x = alcohol, y = chlorides, color = 'Low Quality'),
              data = lowquality, se = FALSE) +
  coord_cartesian(ylim = c(0.0125, 0.15))
# Correlation between Alcohol and Chlorides in High Quality Wine
print('Correlation between Chlorides and Alcohol in High Quality Wine')
cor(highquality$chlorides, highquality$alcohol)
# Correlation between Alcohol and Chlorides in Low Quality Wine
print('Correlation between Chlorides and Alcohol in Low Quality Wine')
cor(lowquality$chlorides, lowquality$alcohol)
```

> Alcohol and Chlorides continue to show themselves as the strongest
pairs of factors in discerning quality level. The curve representing the
high quality wine becomes more horizontal, strengthening the idea that there
is a strict range for the variables in high quality wine.

```{r echo=FALSE, message=FALSE, warning=FALSE, Alcohol_vs_TSD_and_Sugar}
# Alcohol vs Total Sulfur Dioxide vs Quality
# Alpha levels set based on sample sizes (low quality is 33% of total wine)
# This allows for a more accurate visual queue in the plots
al.tsd <- ggplot() +
  geom_point(aes(x = alcohol, y = total.sulfur.dioxide, color = 'High Quality'), 
             data = highquality, alpha = 0.33) +
  geom_point(aes(x = alcohol, y = total.sulfur.dioxide, color = 'Low Quality'), 
             data = lowquality, alpha = 0.21) +
  scale_color_manual("",
                     values = c("High Quality"="darkorange4", 
                                "Low Quality"="blue")) +
  geom_smooth(aes(x = alcohol, y = total.sulfur.dioxide, 
                  color = 'High Quality'),
              data = highquality, se = FALSE) +
  geom_smooth(aes(x = alcohol, y = total.sulfur.dioxide, 
                  color = 'Low Quality'),
              data = lowquality, se = FALSE) +
  coord_cartesian(ylim = c(0, 0.4))
# Alcohol vs Residual Sugar vs Quality
al.rs <- ggplot() +
  geom_point(aes(x = alcohol, y = residual.sugar, color = 'High Quality'), 
             data = highquality, alpha = 0.33) +
  geom_point(aes(x = alcohol, y = residual.sugar, color = 'Low Quality'), 
             data = lowquality, alpha = 0.21) +
  scale_color_manual("",
                     values = c("High Quality"="darkorange4", 
                                "Low Quality"="blue")) +
  geom_smooth(aes(x = alcohol, y = residual.sugar, 
                  color = 'High Quality'),
              data = highquality, se = FALSE) +
  geom_smooth(aes(x = alcohol, y = residual.sugar, 
                  color = 'Low Quality'),
              data = lowquality, se = FALSE)

# Total Sulfur Dioxide Plots and Correlation
print('Total Sulfur Dioxide')
al.tsd
print('Correlation of Total Sulfur Dioxide and Alcohol in High Quality Wine')
cor(highquality$total.sulfur.dioxide, highquality$alcohol)
print('Correlation of Total Sulfur Dioxide and Alcohol in Low Quality Wine')
cor(lowquality$total.sulfur.dioxide, lowquality$alcohol)
# Residual Sugar Plots, Summary and Correlation
print('Residual Sugar')
al.rs
print('Correlation between Sugar and Alcohol in High Quality Wine')
cor(highquality$residual.sugar, highquality$alcohol)
print('Correlation between Sugar and Alcohol in Low Quality Wine')
cor(lowquality$residual.sugar, lowquality$alcohol)
```

> The plot for total sulfur dioxide vs alcohol shows the change between quality
levels in a similar way that we saw with chlorides. High quality wines once 
again have a smaller range for their observed values. For the plot with sugar
we see that the curves representing quality are quite similar. I think this
tells us more about the relationship between sugar and alcohol than it does
with quality. 

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

> The relationship I've observed between alcohol and chloride content seem
to be the strongest pairing of variables. Chloride levels in high quality
wine tend to stay within a certain low range. When combined with a high
alcohol content they are generally seen within high quality wines. 

### Were there any interesting or surprising interactions between features?

> One interesting thing I saw was the alcohol vs sugar plot. The volatility 
between both high and low quality wines is interesting. I was unsure
whether this pattern was due to sugar not being a useful variable, or the fact
that sugar was the most skewed of all observed variables. 

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}
ggplot() +
  geom_boxplot(aes(x = "3 - High", y = alcohol), # High Quality Box Plot
               data = highquality, color = 'darkred') + 
  geom_jitter(aes(x = '3 - High', y = alcohol), # High Quality Jitter
              data = highquality, alpha = 0.1) + 
  geom_boxplot(aes(x = "2 - All", y = alcohol), # All Data Box Plot
               data = wwq, color = 'darkred') +
  geom_jitter(aes(x = '2 - All', y = alcohol), # All Data Jitter
              data = wwq, alpha = 0.1) + 
  geom_boxplot(aes(x = "1 - Low", y = alcohol), # Low Quality Box Plot
               data = lowquality, color = 'darkred') +
  geom_jitter(aes(x = '1 - Low', y = alcohol), # Low Quality Jitter
              data = lowquality, alpha = 0.1) + 
  xlab('Quality Rating (Low = 3 - 5, High = 7 - 9)') +
  ylab('Alcohol (% by Volume)') +
  ggtitle('Alcohol Content in Different Quality Wine')
```

### Description One

> This was the first instance found of a variable that affects quality. Alcohol
is definitely one of the more useful variables in determining quality as well. 
The visual difference in the distribution makes this plot very easy to read 
and understand.

### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
ch.b <- ggplot() +
  geom_boxplot(aes(x = "3 - High", y = chlorides), # High Quality Box Plot
               data = highquality, color = 'darkred') + 
  geom_jitter(aes(x = '3 - High', y = chlorides), # High Quality Jitter
              data = highquality, alpha = 0.05) + 
  geom_boxplot(aes(x = "2 - All", y = chlorides), # All Data Box Plot
               data = wwq, color = 'darkred') + 
  geom_jitter(aes(x = '2 - All', y = chlorides), # All Data Jitter
              data = wwq, alpha = 0.05) + 
  geom_boxplot(aes(x = "1 - Low", y = chlorides), # Low Quality Box Plot
               data = lowquality, color = 'darkred') + 
  geom_jitter(aes(x = '1 - Low', y = chlorides), # Low Quality Jitter
              data = lowquality, alpha = 0.05) + 
  coord_cartesian(ylim = c(0, 0.1)) + 
  xlab('Quality Rating (Low = 3 - 5, High = 7 - 9)') +
  ylab('Chlorides (g / dm^3)') +
  ggtitle('Chlorides in Different Quality Wine')

al.ch <- ggplot(aes(x = alcohol, y = chlorides), data = wwq) +
  geom_point(alpha = 0.1) +
  coord_cartesian(ylim = c(0, 0.1)) +
  geom_smooth(method = 'lm', se = FALSE) +
  xlab('Alcohol (% by Volume)') +
  ylab('Chlorides (g / dm^3)') +
  ggtitle('Chlorides vs Alcohol')

grid.arrange(ch.b, al.ch, ncol = 2)
```

### Description Two

> These plots showed how useful chloride count is as well in determining 
quality. Chloride count being correlated to Alcohol content also proved to be
meaningful when going further in the analysis. 

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
ggplot() +
  geom_point(aes(x = alcohol, y = chlorides, color = 'High Quality'), 
             data = highquality, alpha = 0.33) +
  geom_point(aes(x = alcohol, y = chlorides, color = 'Low Quality'), 
             data = lowquality, alpha = 0.21) +
  scale_color_manual("",
                     values = c("High Quality"="darkorange4", 
                                "Low Quality"="blue")) +
  geom_smooth(aes(x = alcohol, y = chlorides, color = 'High Quality'),
              data = highquality, se = FALSE) +
  geom_smooth(aes(x = alcohol, y = chlorides, color = 'Low Quality'),
              data = lowquality, se = FALSE) +
  coord_cartesian(ylim = c(0.0125, 0.15)) +
  xlab('Alcohol (% by Volume)') +
  ylab('Chlorides (g / dm^3)') +
  ggtitle('Chlorides vs Alcohol in High and Low Quality Wine')
```

### Description Three

> This final plot shows a definite difference in values of chlorides and 
alcohol and their affect on quality. All through out the analysis there
were times where it seemed like high quality wines had more refined values in
the observed variables. This plot is another good example of that.

------

# Reflection 2

> I'm happy to have redone this project. I was happy with the work I did on the
first submission but I was aware it needed some work. The feedback I received
from the marker was very beneficial and helped to change some plots and
analysis. I ended up changing much more (pretty much all of it) than I set
out to do but I think there has been an overall improvement across all levels
of analyses. 

# Reflection 1

> Getting through this project was a struggle to start but I'm happy to have
finished. I tried exploring the data in a lot of different ways to start, 
really trying to find something unique. I got back on track once I started
treating it more like a project to learn R with and to practice analyzing
data. There are other relationships I would continue to look into if I was
to continue with this project. I think I'll realized early into the project
how many different possible ways there is to explore data and I was a little
discouraged. But going through the basics was enough to draw some general
conclusions. 

> I really like how simple it is in R to explore so many different patterns.
Once I get more familiar writing R and using some packages I'm excited to see
how quickly I can find unique patterns in new data. Something I would have 
liked to explore more but didn't was the high quality wines having strict
ranges on several variables. 

